---
title: "PA1_template"
author: "Seher"
date: "October 12, 2017"
output:  
      html_document:  
        keep_md: yes 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Course 5 Week 2 Assignment

This assignment answers the questions detailed below. The aim is to complete the entire assignment in a single R markdown document that can be processed by knitr and be transformed into an HTML file.

## Loading and preprocessing the data

Before processing the data, we do some housekeeping like set the working directory, unzipping the data.

```{r, echo=TRUE}
setwd("C:/Users/s.fazlioglu/Dropbox/Coursera/Course5_reproducible_research")

# getting necessary packages
library(RCurl)
library(dplyr)
library(lubridate)
library(data.table)
library(ggplot2)

# download the file an place it into the folder
if(!file.exists("data")){dir.create("data")}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl,destfile="./data/Dataset.zip",method="libcurl")

# unzip the file
unzip(zipfile="./data/Dataset.zip", exdir="./data")

# get the list of files
list.files("./data", recursive=TRUE)
# the data
activity<- read.csv("./data/activity.csv")
# check what the data looks like
str(activity)
summary(activity)

# removing the files 
file.remove("./data/Dataset.zip")

```

## What is mean total number of steps taken per day?

Calculating the number of steps taken per day

```{r, echo=TRUE}
TotalByDay <- summarise(group_by(activity,date), steps=sum(steps))
```

Make a histogram of the total number of steps taken each day

```{r, echo=TRUE}
hist1 <- ggplot(data = TotalByDay, aes(steps)) + 
    geom_histogram(binwidth = 1500, colour = "white") +
    xlab("Total Number of Steps Taken Each Day") +
    ylab("Count") +
    ggtitle("Histogram of the Total Number of Steps Taken Each Day")
print(hist1)

```

Calculate and report the mean and median of the total number of steps taken per day
 
```{r, echo=TRUE}
mean(na.omit(TotalByDay$steps))

median(na.omit(TotalByDay$steps))
```

## What is the average daily activity pattern?

Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
act_int <- aggregate(steps ~ interval, data = activity, FUN =mean)

ggplot(act_int, aes(interval, steps)) + geom_line()

```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
act_int[which.max(act_int$steps),]
```

## Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
colSums(is.na(activity))
```

Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

```{r}
replace_with_mean <- function(num) replace(num, is.na(num), mean(num, na.rm = TRUE))
meanday <- (activity %>% group_by(interval) %>% mutate(steps = replace_with_mean(steps)))
head(meanday)
```

Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
new_act<-as.data.frame(meanday)
```

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
newTByDay <- summarise(group_by(new_act,date), steps=sum(steps))

hist2 <- ggplot(data =newTByDay , aes(steps)) + 
    geom_histogram(binwidth = 1500, colour = "white") +
    xlab("Total Number of Steps Taken Each Day") +
    ylab("Count") +
    ggtitle("Histogram of the Total Number of Steps Taken Each Day")
print(hist2)

mean(newTByDay$steps)
median(newTByDay$steps)

```

Now compare two results

```{r}
library(grid)
library(gridExtra)

grid.arrange(hist1, hist2, ncol = 2)

```

Compare the mean and median

```{r}
mean(na.omit(TotalByDay$steps))
median(na.omit(TotalByDay$steps))

mean(newTByDay$steps)
median(newTByDay$steps)
```

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
new_act$Weekdays <- ifelse(weekdays(as.Date(new_act$date)) %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"), "Weekday", "Weekend")
head(new_act)
```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r}
new_act <- (new_act %>% group_by(interval, Weekdays) %>% summarise(Mean = mean(steps)))
ggplot(new_act, mapping = aes(x = interval, y = Mean)) + geom_line() +
    facet_grid(Weekdays ~.) + xlab("Interval") + ylab("Mean of Steps") +
    ggtitle("Comparison of Average Number of Steps in Each Interval")
```

